RDF Validation of Metadata \\ on Highly-Complex Person-Level Data

### 2015-02-06 paper/paper7/paper7-appendix.tex
disco validation rules

\subsection{Class-Specific Property Ranges}

{\em Class-Specific Property Range} restricts the range of object and data properties for individuals within a specific context (e.g. class, shape, application profile).
The values of each member property of a class may be limited by their value type, such as xsd:string or foaf:Person\footnote{{\em R-29-CLASS-SPECIFIC-RANGE-OF-RDF-OBJECTS} and {\em R-36-CLASS-SPECIFIC-RANGE-OF-RDF-LITERALS}}. 
\textbf{{\em DISCO-C-XX:}} 
Only {\em disco:Question}s can have {\em disco:questionText} relationships to literals of the datatype {\em rdf:langString}.

\begin{DL}
$\neg$Question $\sqsubseteq$ $\neg\exists$ questionText.langString
\end{DL}

@@

Resources can only be members of listed controlled vocabularies\footnote{{\em R-32-MEMBERSHIP-OF-RDF-OBJECTS-IN-CONTROLLED-VOCABULARIES} and
{\em R-39-MEMBERSHIP-OF-RDF-LITERALS-IN-CONTROLLED-VOCABULARIES}}.

\textbf{{\em DISCO-C-XX}}.
{\em disco:SummaryStatistics} can only have {\em disco:summaryStatisticType} relationships to {\em skos:Concept}s which must be members of the controlled vocabulary {\em ddicv:SummaryStatisticType} which is a {\em skos:ConceptScheme}.

\begin{DL}
SummaryStatistics $\sqsubseteq$ \\
$\forall summaryStatisticType.Concept \sqcap \forall inScheme . A$ \\
$A \equiv ConceptScheme \sqcap \{SummaryStatisticType\}$
\end{DL}

@@

Depending on the property semantics,
there are cases where two different literal values must have
a specific ordering with respect to an operator. 
P1 and P2 are the datatype properties we need to compare and 
OP is the comparison operator (\textless, \textless=, \textgreater, \textgreater=, =, !=)\footnote{{\em R-43-LITERAL-VALUE-COMPARISON}}.
The {\em COMP Pattern}, one of the Data Quality Test Pattern, can be used to validate the {\em Literal Value Comparison} constraint \cite{Kontokostas2014}:

@@

\textbf{{\em DISCO-C-XX}}.
{\em disco:startDate}s must be before (‘\textless’) the {\em disco:endDate}s.
To validate this constraint we bind the variables as follows (P1: {\em disco:startDate}, P2: {\em disco:endDate}, OP: \textless).

@@

Examples for {\em Mathematical Operations} are the addition of two dates, the addition of days to a start date, and statistical computations (e.g. average, mean, sum).
\textbf{{\em DISCO-C-XX:}}
The sum of {\em disco:percentage} {\em xsd:double} values of all codes (represented as {\em skos:Concept}s) of a code list ({\em skos:ConceptScheme} or {\em skos:OrderedCollection}), serving as representation of a particular {\em disco:Variable}, must exactly be 100  
(not expressible in DL)\footnote{{\em R-42-MATHEMATICAL-OPERATIONS} and {\em R-41-STATISTICAL-COMPUTATIONS}}.

Default values for objects and literals are inferred automatically.
It should be possible to declare the default value for a given property, e.g. so that input forms can be pre-populated and to insert a required property that is missing in a web service call\footnote{{\em R-31-DEFAULT-VALUES-OF-RDF-OBJECTS} and {\em R-38-DEFAULT-VALUES-OF-RDF-LITERALS}}.
\textbf{{\em DISCO-C-XX:}}
The value 'true' for the property {\em disco:isPublic} ({\em xsd:boolean}) indicates that the data set ({\em disco:LogicalDataSet}) can be accessed (usually downloaded) by anyone.
Per default, access to data sets should be restricted ('false').

@@

For particular data properties, values must be stated for predefined languages (not expressible in DL)\footnote{{\em R-47-LANGUAGE-TAG-MATCHING}}.
\textbf{{\em DISCO-C-XX:}}
There must be an English variable name ({\em skos:notation}) for each {\em disco:Variable} within {\em disco:LogicalDataSet}s.

@@

@@

@@

@@

In DDI, variables, questions, and categories are typically organized in a particular order. 
For obtaining this order, {\em skos:OrderedCollection} resources are used. 
A collection of variables, e.g, is represented as being of the type {\em skos:OrderedCollection} containing multiple variables (each represented as {\em skos:Concept}) in a {\em skos:memberList}.

>>>>

\subsection{Class-Specific Property Range*}

{\em Class-Specific Property Range}\footnote{{\em R-29-CLASS-SPECIFIC-RANGE-OF-RDF-OBJECTS}, {\em R-36-CLASS-SPECIFIC-RANGE-OF-RDF-LITERALS}} restricts the range of object and data properties for individuals within a specific context (e.g. class, shape, application profile).
The values of each member property of a class may be limited by their value type, such as \emph{xsd:string} or \emph{foaf:Person}.

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX:}} 
Only {\em disco:Question}s can have {\em disco:questionText} relationships to literals of the datatype {\em rdf:langString}.

\begin{DL}
$\neg$Question $\sqsubseteq$ $\neg\exists$ questionText.langString
\end{DL}
\end{itemize}

\subsection{Class-Specific Reflexive Object Properties*}

Using DL terminology \emph{Class-Specific Reflexive Object Properties} is called local reflexivity - a set of individuals (of a specific class) that are related to themselves via a given role \cite{Kroetzsch2012}.

@@

Resources can only be members of listed controlled vocabularies\footnote{{\em R-32-MEMBERSHIP-OF-RDF-OBJECTS-IN-CONTROLLED-VOCABULARIES}, 
{\em R-39-MEMBERSHIP-OF-RDF-LITERALS-IN-CONTROLLED-VOCABULARIES}}.

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX}}:
{\em disco:SummaryStatistics} can only have {\em disco:summaryStatisticType} relationships to {\em skos:Concept}s which must be members of the controlled vocabulary {\em ddicv:SummaryStatisticType} which is a {\em skos:ConceptScheme}.

\begin{DL}
SummaryStatistics $\sqsubseteq$ $\forall summaryStatisticType.A$ \\
$A \equiv Concept \sqcap \forall inScheme . B$ \\
$B \equiv ConceptScheme \sqcap \{SummaryStatisticType\}$
\end{DL}
\end{itemize}

\subsection{IRI Pattern Matching}

IRI pattern matching applied on subjects, properties, and objects\footnote{\emph{R-21-IRI-PATTERN-MATCHING-ON-RDF-SUBJECTS}, \emph{R-22-IRI-PATTERN-MATCHING-ON-RDF-OBJECTS}, \emph{R-23-IRI-PATTERN-MATCHING-ON-RDF-PROPERTIES}}.

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX}}: \emph{disco:Study} resources must match a given IRI pattern.
\end{itemize}

@@

Depending on the property semantics,
there are cases where two different literal values must have
a specific ordering with respect to an operator. 
\emph{P1} and \emph{P2} are the datatype properties we need to compare and 
\emph{OP} is the comparison operator (\textless, \textless=, \textgreater, \textgreater=, =, !=)\footnote{{\em R-43-LITERAL-VALUE-COMPARISON}}.
The {\em COMP Pattern}, one of the Data Quality Test Patterns, can be used to validate the {\em Literal Value Comparison} constraint \cite{Kontokostas2014}:

@@

\begin{itemize}
	\item \textbf{{\em DISCO-C-XX}}:
{\em disco:startDate}s must be before (‘\textless’) {\em disco:endDate}s.
To validate this constraint we bind the variables as follows (P1: {\em disco:startDate}, P2: {\em disco:endDate}, OP: \textless). 
\end{itemize}

With this constraint objects of object properties can be ordered as well as literals of data properties and objects of object properties\footnote{\emph{R-121-SPECIFY-ORDER-OF-RDF-RESOURCES}, \emph{R-217-DEFINE-ORDER-FOR-FORMS/DISPLAY}}.

In DDI, variables, questions, and codes/categories are typically organized in a particular order. 
For obtaining this order, {\em skos:OrderedCollection} resources are used.

\begin{itemize}
	\item \textbf{{\em DISCO-C-ORDERING-01}}: If \emph{disco:Variable}s of a given \emph{disco:LogicalDataSet} should be ordered, a collection of variables must be present in the data and connected with the data set. The collection of variables is of the type {\em skos:OrderedCollection} containing multiple variables (each represented as {\em skos:Concept}) in a {\em skos:memberList}. 
	\item \textbf{{\em DISCO-C-ORDERING-02}}: If \emph{disco:Question}s of a given \emph{disco:Questionnaire} should be ordered, a collection of questions must be present in the data and connected with the questionnaire. The collection of questions is of the type {\em skos:OrderedCollection} containing multiple questions (each represented as {\em skos:Concept}) in a {\em skos:memberList}. 
	\item \textbf{{\em DISCO-C-ORDERING-03}}: If codes/categories (\emph{skos:Concept}s) of a given \emph{disco:Representation} of a given \emph{disco:Variable} should be ordered, the variable representation should also be of the type {\em skos:OrderedCollection} containing multiple codes/categories (each represented as {\em skos:Concept}) in a {\em skos:memberList}. 
\end{itemize}

\section{Validation Levels}

Different levels of severity (priority)\footnote{\emph{R-205-VARYING-LEVELS-OF-ERROR}, \emph{R-135-CONSTRAINT-LEVELS}, \emph{R-158-SEVERITY-LEVELS-OF-CONSTRAINT-VIOLATIONS}, \emph{R-193-MULTIPLE-CONSTRAINT-VALIDATION-EXECUTION-LEVELS}} should be assigned to constraints.
Possible validation levels could be: informational, warning, error, fail, should, recommended, must, may, optional, closed (only this) constraints, open (at least this) constraint.

For \emph{Disco} each constraint should be assigned to exactly one \emph{validation level}.

\section{String Operations}

Many different \emph{string operations}\footnote{\emph{R-194-PROVIDE-STRING-FUNCTIONS-FOR-RDF-LITERALS}} are possible.
Some constraints require building new strings out of other strings.
Calculating the string length would also be another constraint of this type.

\begin{itemize}
	\item \textbf{{\em DISCO-C-STRING-OPERATIONS-01}}: The title of a study (\emph{dcterms:title}) (e.g. 'EU-SILC 2005') may be calculated out of the title of the containing series (\emph{dcterms:title}) (e.g. 'EU-SILC') and the human-readable label of the study (\emph{rdfs:label}) (e.g. '2005').   
\end{itemize}

\subsection{Context-Specific Valid Classes}

What types are valid in a specific context?\footnote{\emph{R-209-VALID-CLASSES}} 
Context can be an input stream, a data creation function, or an API.

\begin{itemize}
	\item \textbf{{\em DISCO-C-CONTEXT-SPECIFIC-VALID-CLASSES-01}}: For future versions of \emph{Disco}, out-dated classes can be marked as deprecated.
\end{itemize}

\subsection{Context-Specific Valid Properties}

What properties can be used within this context?\footnote{\emph{R-210-VALID-PROPERTIES}} 
Context can be an data receipt function, data creation function, or API.

\begin{itemize}
	\item \textbf{{\em DISCO-C-CONTEXT-SPECIFIC-VALID-PROPERTIES-01}}: For future versions of \emph{Disco}, out-dated properties can be marked as deprecated.
\end{itemize}

\emph{Default values}\footnote{{\em R-31-DEFAULT-VALUES-OF-RDF-OBJECTS}, {\em R-38-DEFAULT-VALUES-OF-RDF-LITERALS}} for objects and literals are inferred automatically.
It should be possible to declare the default value for a given property, e.g. so that input forms can be pre-populated and to insert a required property that is missing in a web service call.

\begin{itemize}
	\item \textbf{{\em DISCO-C-DEFAULT-VALUES-01:}}
The value 'true' for the property {\em disco:isPublic} ({\em xsd:boolean}) indicates that the data set ({\em disco:LogicalDataSet}) can be accessed (usually downloaded) by anyone.
Per default, access to data sets should be restricted ('false').
\end{itemize}

@@

Examples for {\em Mathematical Operations}\footnote{{\em R-42-MATHEMATICAL-OPERATIONS}, {\em R-41-STATISTICAL-COMPUTATIONS}} are the addition of two dates, the addition of days to a start date, and statistical computations (e.g. average, mean, sum).

\begin{itemize}
	\item \textbf{{\em DISCO-C-MATHEMATICAL-OPERATIONS-01:}}
The sum of {\em disco:percentage} (datatype: {\em xsd:double}) values of all codes (represented as {\em skos:Concept}s) of a code list ({\em skos:ConceptScheme} or {\em skos:OrderedCollection}), serving as representation of a particular {\em disco:Variable}, must exactly be 100.
\end{itemize}

@@

For particular data properties, values must be stated for predefined languages\footnote{{\em R-47-LANGUAGE-TAG-MATCHING}}.

\begin{itemize}
	\item \textbf{{\em DISCO-C-LANGUAGE-TAG-MATCHING:}}
There must be an English variable name ({\em skos:notation}) for each {\em disco:Variable} within {\em disco:LogicalDataSet}s.
\end{itemize}

@@

\subsection{Whitespace Handling}

\subsection{HTML Handling}

@@

\subsection{Handle RDF Collections}

@@

\subsection{Exploiting Class/Property Specialization Ontology Axioms}

\subsection{Cardinality Shortcuts*}

\subsection{Aggregation}

@@
